Neural Connectome of the Ctenophore Statocyst

### 2025-05-06 manuscript/Manuscript.qmd
MS

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig; @Hernandez-Nicaise; @Jager]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@Moroz]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior. Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@Horiddge1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@Horridge1965; @Tamm1982; @Tamm2014]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

@@

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants.

@@

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013; @Moroz2014]. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024]. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015;@roberts2022].

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances \[[@waggett1999; @colin2010; @courtney2020]\]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

>>>>

Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates. These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column. Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system. Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011]. These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022]. Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018]. It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022]. In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015]. This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems. However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior. Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.

To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ. This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980]. Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011]. A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @Tamm_2014b]. Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997]. In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997]. Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997]. These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].

@@

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999]. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999]. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999]. We assigned each of the 909 skeletonised cells to one of the four quadrants.

@@

Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics. This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns. Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs. inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014]. However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.

Strikingly, this organization resembles circuit motifs found in other animals. For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024]. Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015; @roberts2022].

@@

How do differences in balancer cilia coordination across anatomical planes shape behavior? In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020]. Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.
